

```{r, echo=FALSE, message=FALSE}
library(chron) #helpful reference: #http://www.stat.berkeley.edu/~s133/dates.html
library(ggplot2)
library(stringr)
library(gdata)
library(plyr)
```

Import the race results (parsed from HTML via python's BeautifulSoup), corral assignments, and the time windows associated with each corral.

```{r}
setwd("/Users/clmcleod/GitHub/sideProjects/Running/RunningData/PhillyMarathon2014/")
fullResults <- read.csv("Philly2014RaceResults-HTML/fullResults.csv")

corralAssignments <- read.csv("corralAssignments.csv")

predict <- read.csv("corralPredictions.csv")

predict[, 2:5] <- lapply(predict[,2:5], 
                         function(x){as.POSIXct(strptime(x, '%H:%M:%S'))} )
```


Clean up the data. Two individuals have NA bib numbers due to legacy quaalification.
```{r, warning = FALSE}
#remove leading spaces
corralAssignments <- trim(corralAssignments)
corralAssignments$registration <- gsub("Half Marathon", "Half", corralAssignments$registration)

# remove 8k results, since we aren't interested in those
c<-subset(corralAssignments, registration %in% c("Half", "Marathon"))
```

Now, match the registrant and finisher data by merging across bib number. Make sure the demographic information matches across the two tables - it doesn't always, and we'll remove these results.
```{r, message=FALSE, warning=FALSE}
#match up registration with results info by bib #
#the NAs introduced by coercion here are the 2 legacy runners
fullResults$match<-as.integer(fullResults$bib)
c$match<-as.numeric( levels(c$bib) )[c$bib]
m <- merge(fullResults, c, by.x = "match", by.y = "match", sort=TRUE)

#flag observation as suspicious if sex and age aren't consistent b/w data sources
m$flag <- ( m$sex.x == m$sex.y  &  abs(m$age.x-m$age.y)<2 )
suspicious <- m[!m$flag, ]

#clean is a data frame that contains both half and full marathon data, for finishers and non-finishers
#we've removed the legacy runners and the registrants who had inconsistent demographic information
clean<-m[m$flag, ]
```

Next, convert the HH:MM:SS times to just seconds for easier modeling later. These functions were taken directly from http://stackoverflow.com/questions/10835908/is-there-a-way-to-convert-mmss-00-to-seconds-00-in-r

```{r, echo=FALSE}
toSeconds <- function(x){
   if (!is.character(x)) stop("x must be a character string of the form H:M:S")
   if (length(x)<=0) return(x)

   unlist(
      lapply(x,
         function(i){
            i <- as.numeric(strsplit(i,':',fixed=TRUE)[[1]])
            if (length(i) == 3) 
               i[1]*3600 + i[2]*60 + i[3]
            else if (length(i) == 2) 
               i[1]*60 + i[2]
            else if (length(i) == 1) 
               i[1]
            else if (length(i) == 0)
                NA
         }  
      )  
   )  
} 

secondsToString <- function(x,digits=2){
   unlist(
      lapply(x,
         function(i){
            # fractional seconds
            fs <- as.integer(round((i - round(i))*(10^digits)))
            fmt <- ''
            if (i >= 3600)
               fmt <- '%H:%M:%S'
            else if (i >= 60)
            fmt <- '%M:%S'
            else
               fmt <- '%OS'

            i <- format(as.POSIXct(strptime("0:0:0","%H:%M:%S")) + i, format=fmt)
            if (fs > 0)
               sub('[0]+$','',paste(i,fs,sep='.'))
            else
               i
         }
      )
   )
}
```

```{r}
clean$netTime<- as.POSIXct(strptime(clean$net, '%H:%M:%S'))
clean$netSec <- toSeconds(as.character(clean$net))
```

Now, we'll create a designation for registrants who didn't finish (DNF). Now we have all the pieces necessary to create a graph showing finisher outcomes. We can also tabulate the data directly.
```{r}
# if a participant had a net time of "", change event to DNF
# (need to add DNF as a level first)
levels(clean$event) <- c(levels(clean$event), "DNF")
clean$event[clean$net == ""] = "DNF"
clean$DNF = ifelse(clean$net == "", 1, 0)

#bar plot of registration-finisher data
ggplot(clean)+
  geom_bar(aes(x=registration, stat="bin", fill=event))+
  ylab("# Participants")+
  xlab("Registered Event")+
  guides(fill=guide_legend(title="Actually Ran"))

clean$reg.ran <- interaction(clean$registration, clean$event)
summary(clean$reg.ran)

#remove some duplicate fields from the dataset
drops <- c("bib.x", "sex.x", "age.x", "match", "flag")
clean <- clean[, !names(clean) %in% drops]
clean <- rename(clean, c("sex.y" = "sex", "age.y"="age", "bib.y"="bib"))
```

I think a much better visualization for this is a Sankey diagram. I used a great online tool called Sankeymatic - here's the input for that.
```{r}
# http://sankeymatic.com/
# 
# Marathon Registrants[10048] Marathon Finishers #8904B1
# Marathon Registrants[1172] Half Marathon Finishers #8904B1
# Marathon Registrants [2797] DNF #8904B1
# Marathon Registrants [2] 8k Finishers
# 
# Half Marathon Registrants [11228] Half Marathon Finishers #04B431
# Half Marathon Registrants [2795] DNF #04B431
# Half Marathon Registrants [28] 8k Finishers #04B431
# 
# 8k Registrants [1670] 8k Finishers #045FB4
# 8k Registrants [644] DNF #045FB4
# 
# :DNF #848484
# :Half Marathon Finishers #04B431
# :Half Marathon Registrants #04B431
# :Marathon Finishers #8904B1
# :Marathon Registrants #8904B1
```

Finally, some quick linear modeling ... 

```{r}
#r.mar contains just data for marathon finishers
r.mar <- clean[ clean$reg.ran %in% c("Marathon.Marathon"), ]

summary(lm( netSec ~  sex * age, data = r.mar))

summary(lm( netSec ~  corral, data = r.mar))

# does demographic information predict who finishes?
marathon.registrants <- clean[ clean$registration %in% c("Marathon"),]
summary(glm(DNF ~ sex * age, family = "binomial", data = marathon.registrants))
```

... and some graphs of marathon finish time as a function of age and gender:

```{r}


ggplot(r.mar)+
  geom_density2d(aes(x=age, y= netSec, color=sex) ) +
  geom_point(aes(x=age, y=netSec), alpha = 0.02, size = 3) +
  facet_grid(.~sex)+
  ylab("Time (seconds)")+
  xlab("Age")

ggplot(data = r.mar, aes(x=age, y= netSec)) +
  stat_density2d(geom="tile", aes(fill = ..density..), contour = FALSE)+
  facet_grid(.~sex)+
  ylab("Time (seconds)")+
  xlab("Age")
```


